---
title: "01 - R@McMasterU: R Basics + GitHub"
author: "Ahmad Sofi-Mahmudi"
date: "`r Sys.Date()`"
output: html_document
---

# Purpose
In the first session, we explore the very basic functions of R for biostatistics and also have a glance over GitHub.

# Installing and loading packages
You can install packages using __*install.packages()*__ function. The important thing is that you should put the package name into quotation marks:
```{r, eval=FALSE}
install.packages("tidyverse")
```

To install multiple packages, you should use concatenate function __*c()*__:
```{r, eval=FALSE}
install.packages(c("tidyverse", "metafor", "FactoMineR"))
```

Another way is to use RStudio graphical interface: **Tools -> Install Packages**.

After installing, to use the package, you should load it using __*library()*__ function:
```{r, eval=FALSE}
library(tidyverse)
```

Note that loading a package doesn't require quotation marks.

# Getting and setting the working directory
The first step of almost every project is to set the working directory in order to have reproducible workflow.

To *get* the current working directory (*wd*), we use __*getwd()*__ function:
```{r}
getwd()
```

To *set* our desired wd, we use __*setwd()*__. In this function, we should specify the path, for example:
```{r}
setwd("/home/ahmad/Documents/GitHub/R-McMasterU/01 - R Basics + GitHub/")
```

However, using *setwd* has some drawbacks which makes it irreproducible. For example, if you want to use the code on some other computer or change directory names, you could face some difficulties. These problems can be solved by making an *rproj* file but not everyone uses RStudio. Another way that I recommend is to use __*here()*__ function. You should download and install it first:
```{r, eval=FALSE}
install.packages("here")
library(here)
```

```{r, echo=FALSE}
library(here)
```


And then by using the __*here()*__ function, the working directory will be set to the directory where our *.R* or *.Rmd* file is located.
```{r}
here()
```

You can also use it inside other functions:
```{r, eval=FALSE}
here("one", "two", "McMaster.R")
```


You also can use just a particular function from a package:
```{r, eval=FALSE}
dplyr::filter()
```

There are some built-in datasets in R. You can access to the list with __*data()*__ function:
```{r}
data()
```

And to load a built-in dataset, we use the same function with the dataset name in it:
```{r}
data(mtcars)
```

# Data types
There are 6 data types in R:
- character: "a", "swc"
- numeric: 2, 15.5
- integer: 2L (the L tells R to store this as an integer)
- logical: TRUE, FALSE
- complex: 1+4i (complex numbers with real and imaginary parts)

We use __*typeof()*__ function to know the object's data type:
```{r}
typeof(T)
```

__*class()*__ function can also be used to determine the high-level type of object. We also have __*mode()*__ and __*storage.mode()*__ functions. To know more about the differences, please take a look at the following codes:
```{r}
x <- 1L
print(c(class(x), mode(x), storage.mode(x), typeof(x)))

x <- 1
print(c(class(x), mode(x), storage.mode(x), typeof(x)))

x <- letters
print(c(class(x), mode(x), storage.mode(x), typeof(x)))

x <- TRUE
print(c(class(x), mode(x), storage.mode(x), typeof(x)))

x <- cars
print(c(class(x), mode(x), storage.mode(x), typeof(x)))

x <- cars[1]
print(c(class(x), mode(x), storage.mode(x), typeof(x)))

x <- cars[[1]]
print(c(class(x), mode(x), storage.mode(x), typeof(x)))

x <- matrix(cars)
print(c(class(x), mode(x), storage.mode(x), typeof(x)))

x <- new.env()
print(c(class(x), mode(x), storage.mode(x), typeof(x)))

x <- expression(1 + 1)
print(c(class(x), mode(x), storage.mode(x), typeof(x)))

x <- quote(y <- 1 + 1)
print(c(class(x), mode(x), storage.mode(x), typeof(x)))

x <- ls
print(c(class(x), mode(x), storage.mode(x), typeof(x)))
```

## Data structures in R
- atomic vector
- list
- matrix
- data frame
- factors

## Converting data types
To convert data types, we use these functions:
- __*as.logical()*__
- __*as.numeric()*__
- __*as.character()*__
- __*as.factor()*__


# Reading data
## .csv files
R has a built-in function for reading .csv files. __*read.csv()*__ is a wrapper function for __*read.table()*__ that mandates a comma as separator and uses the input file's first line as header that specifies the table's column names.

An example:
```{r, eval=FALSE}
csv_file = read.csv(file = "path_in_quotation",
                    header = FALSE,
                    sep = "", 
                    quote = "\"'",
                    na.strings = "NA",
                    stringsAsFactors = FALSE,
                    encoding = "unknown")
```

## .xlsx files
I prefer to use __*read.xlsx()*__ function. You should install and load *openxlsx* package first:
```{r, eval=FALSE}
install.packages("openxlsx")
library(openxlsx)
```

And now read the .xlsx file:
```{r, eval=FALSE}
xlsx_file = read.xlsx(xlsxFile = "path_in_quotation",
                      sheet = sheetnumber,
                      startRow = 1,
                      na.strings = "NA")
```


## .dta files
.dta is the file format that STATA uses. To read it, we need another package. *haven*, *readstata13*, and *foreign* are all good. I have used *readstata13* and *foreign* for my projects. *foreign* is better in that sense that it has functions to read other formats for popular statistical software such as SAS or SPSS. So, we go with *foreign*.
First, installing and loading it:
```{r, eval=FALSE}
install.packages("foreign")
library(foreign)
```

And then, using __*read.dta*__ to read our file:
```{r, eval=FALSE}
dta_file = read.dta(file = "path_in_quotation")
```

# Writing data
For almost all read.something() function, we have write.thatthing() function as well. We will learn how to use write functions where needed.

# Creating data in native R environment
We also can create data in R environment, almost every data type.
## Vectors
Vectors are one-dimensional data sets and easiest to create. Here are all the ways we can do so:
```{r}
c(1, 2, 3)
2:5
seq(2, 3, by = 0.5)
rep(1:2, times = 3)
rep(1:2, each = 3)
```
## Matrixes
Matrix is a two dimensional data structure. We create a matrix with the __*matrix()*__ function:
```{r}
matrix(data = NA,
       nrow = 1,
       ncol = 1,
       dimnames = NULL)
```


## Lists
A list is collection of elements which can be of different types. We use the __*list()*__ function to create lists:
```{r}
list(x = 1:5,
     y = c("a", "b", "c"),
     z = TRUE,
     s = 1+4i)
```


## Data frames
A special case of a list where all elements are the same length. The function is __*data.frame()*__:
```{r}
data.frame(x = 1:3,
           y = c("a", "b", "c"))
```


# Selecting and subsetting data
We can use *[]* to select and subset data which is a built-in R feature. We also can use __*select()*__ and __*filter()*__ functions from *dplyr* package (comes with *tidyverse* package). Here are examples:

## Selecting vector elements
By position:
```{r}
x = 1:10
# The fourth element
x[4]
# All but the fourth.
x[-4]
# Elements two to four.
x[2:4]
# All elements except two to four.
x[-(2:4)]
# Elements one and five.
x[c(1, 5)]
```

By value:
```{r}
x = -1:10
# Elements which are equal to 10.
x[x == 10]
# All elements less than zero.
x[x < 0]
```

Named vectors:
For this, we first should assign names to our values with __*names()*__ function (pretty much like English, right?):
```{r}
x = c("Ahmad", "Sofi-Mahmudi")
names(x) = c("First", "Last")
# Element with name 'Last'.
x["Last"]
```


## Selecting matrixes and data frames elements
Matrix:
```{r}
x = 1:3
# Create a matrix from x.
m <- matrix(x, nrow = 3, ncol = 3)
# Select a row
m[2, ]
# Select a column
m[, 1]
# Select an element
m[2, 3]
# Transpose
t(m)
# Matrix multiplication
n <- matrix(x, nrow = 3, ncol = 3)
m %*% n
```

Data frame:
```{r}
df <- data.frame(x = 1:3, y = c('a', 'b', 'c'))
# List subsetting
df$x
df[[2]]
# Select second column
df[, 2]
# Select second row
df[2, ]
# Select second column and second row
df[2, 2]
```


## Selecting list elements
```{r}
l <- list(x = 1:5, y = c('a', 'b'))
# Second element of l.
l[[2]]
# New list with only the first element.
l[1]
# Element named x
l$x
# New list with only element named y.
l['y']
```

## Using dplyr
*dplyr* has two functions for this: __*filter()*__ and __*select()*__. __*filter()*__ applies a condition on a variable whereas __*select()*__ selects variables from a data frame. Example:
```{r}
library(dplyr)
# Loading mtcars dataset
data(mtcars)
# Select two variables: mpg, cyl
select(mtcars, c(mpg, cyl))
# Filter rows with mpg > 20 and cyl > 4
filter(mtcars, mpg > 20 & cyl > 4)
```


# Git
Git is a version controlling software. You can use GitHub Desktop (desktop.github.com) but here we take a glance on Git commands.

## Setting up a Git folder
First, changing directory to the one we want to set up Git in:
```{r, eval=FALSE}
cd myproject
```

Then, we use initialize a Git repository using __*init*__ function:
```{r, eval=FALSE}
git init
```

We create a file in the folder called *mcmaster.R*, but Git doesn't track it by default. To track it using Git, we use __*add*__ function:
```{r, eval=FALSE}
git add mcmaster.R
```

We use __*status*__ function to know the overall status of the files in the folder:
```{r, eval=FALSE}
git status
```

A commit is a record of what changes you have made since the last time you made a commit. Essentially, you make changes to your repo (for example, adding a file or modifying one) and then tell git to put those changes into a commit.
We do that using the __*commit*__ function:
```{r, eval=FALSE}
git commit -m "Your message about the commit"
```

Note that you always need to have a title or message for your commit in order to know what have you done since the last commit. This will track your project perfectly and organizes your thoughts regarding what you changed.

## GitHub
Of course you want to use GitHub to work collaboratively in a team! So, using Git locally is not enough. After creating your repository on GitHub website, you use __*remote*__ function to grab that repo. It will automatically create a folder with the name of the repo:
```{r, eval=FALSE}
git remote add origin https://github.com/username/mynewrepository.git
```

Then, after we made changes and committed the changes, we need to __*push*__ it online. To do so, we do the following:
```{r, eval=FALSE}
git push -u origin master
```

